Redundant calibration as a complex optimization problem

### 2017-10-11 paperG2.tex
newest draft from Gianni

\title[Unknown]{Redundant calibration as a complex optimization problem}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{4}$National Radio Astronomy Obs., Socorro NM\\
$^{5}$Cavendish Lab., Cambridge UK\\
$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - ''redundant calibration" - and boosts sensitivity on specific scales. We show that redundant calibration can be formulated in the domain of complex functions of a complex variable. This framework reduces the 

@@

\textbf{[Could we add JS Kenyon and SJ Wijnholds to author list?]}

@@

\section{Introduction}
The effects (i.e. the ionosphere, antenna gains, etc.) a celestial signal experiences along its propagation path are modelled with the radio interferometric measurement equation (RIME) \citep{ME1,RRIME1}.
Calibration is the procedure by which we remove the errors that these propagation effects introduce into interferometric data and is normally accomplished using a 
least-squares solver. In the case where we only model the antenna gains, which is the most basic form of the RIME, we can define calibration as finding the antenna gains that minimize the difference between our observed and predicted visibilities.

There are two recent developments in skymodel-based calibration which should be explicitly mentioned at this point, since they form the foundation this paper builds on. The first is the discovery of \textsc{StEfCal} (Statistically Efficient Calibration) \citep{Mitchell:MWA-cal,Salvini2014}, which 
is an alternating direction implicit (ADI) skymodel-based calibration algorithm. It works by assuming the gains are known while solving for their conjugates and then performing the reverse. 
This procedure results in the linearization of the calibration algorithm and reduces the computational complexity of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$, where $N$ denotes the number of antennas in the array. 
The second development is the idea: instead of breaking the calibration problem 
into its real and imaginary parts and then solving for the real and imaginary parts of the unknown parameters separately we can treat the antenna gains and their 
conjugates as separate parameters to estimate \citep{Smirnov2015}. \citet{Smirnov2015} also showed that the \textsc{StEfCal} algorithm itself is merely a special case of this
gain-and-conjugate framework.

Two baselines are redundant if their baseline difference vectors have the same length and orientation, which imply that they measure the same $uv$-modes. When we use a redundant array layout, we dramatically 
reduce the number of unknowns. If our array is redundant enough we reduce the number of unknowns to such an extent that we are able to solve for the unknown 
observed visibilities themselves in addition to the antenna gains. Redundant calibration is the procedure by which we estimate both the unknown observed visibilities and the antenna gains.

Redundant calibration was popularized in the 1980's. \citet{Noordam1982} used redundant calibration and the WSRT (Westerbork Synthesis Radio Telescope) to produce high-dynamic range images of 3C84.
The first implementations of redundant calibration made use of the logarithm function, because when the logarithm is applied to the scalar measurement equation, redundant 
calibration is transformed into a linear problem \citep{Wieringa1992,Camps2003,Liu2010}. The implementation associated with this approach is commonly referred to as \textsc{logcal}.
We briefly discuss \textsc{logcal} in Appendix~\ref{sec:logcal}. The next improvement in redundant calibration was to perform a first order Taylor expansion, which is equivalent to the Gauss-Newton (GN) algorithm \citep{Kurien2016}, of the measurement equation and then to solve for the
perturbations of the model-parameters \citep{Liu2010}. \citet{Liu2010} perform the first order Taylor expansion of the measurement equation by splitting the unknown parameters into their phase-and-amplitude
components. The implementation of  this approach is called \textsc{lincal}. \textsc{lincal} is derived from first principals in Appendix~\ref{sec:lincal}. 
Currently, redundant calibration is usually implemented using a layered algorithmic strategy. \textsc{logcal} and \textsc{lincal} are both implemented in \textsc{omnical}\footnote{https://github.com/jeffzhen/omnical}, which is a layered redundant calibration pipeline \citep{Zheng2014,Ali2015}.
The solutions of \textsc{logcal} are used as initial input for \textsc{lincal}, since the GN algorithm requires that its starting parameter guesses be in the vicinity 
of the true solution.

It is also interesting to mention here that a completely different calibration strategy can also be used to calibrate redundant arrays, \textsc{CorrCal} \citep{Sievers2017}\footnote{https://github.com/sievers/corrcal2}. In \textsc{CorrCal} the unknown 
visibilities are no longer solved for. The sky is actually modelled, via two covariance matrices. 
The first covariance matrix is associated with known point sources. The second covariance matrix models the remaining sky in a statistical sense. The unknown gains, however, are still 
estimated. We do not elaborate any further on this approach in this paper.

New instruments like LOFAR (Low Frequency Array) \citep{Noorishad2012}, PAPER (Precision Array for Probing the Epoch of Reionization) \citep{Ali2015} and HERA (Hydrogen Epoch of Reionization Array) \citep{deboer2015} have also sparked a renewed interest in redundant calibration. Using redundancy is especially advantageous when we are trying to detect the EoR (Epoch of Reionization) 21 cm signal,
which is inherently quite faint \citep{Parsons2012}. Since redundant baselines make statistically independent measurements of the same $uv$-modes redundancy can greatly 
improve the SNR (signal-to-noise ratio) of EoR detection experiments. This explains why PAPER and HERA are highly redundant arrays, as they are both primarily used for EoR 21 cm detection experiments.

\noindent
\textbf{[GIANNI TO REWRITE AND EXTEND EOR SECTION]}

In this paper we propose to use the complex framework introduced by \citet{Smirnov2015} instead of the traditional amplitude-and-phase approach used by \textsc{lincal}.
We also propose to use the Levenberg-Marquardt (LM) gradient-based algorithm instead of the Gauss-Newton (GN) algorithm as it exhibits better convergence properties.
Using the LM algorithm reduces the number of algorithmic layers that are needed to perform redundant calibration.

Two ways of speeding up redundant calibration have recently been proposed. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities \citep{Wijnholds2012,Marthi2014}. Note that the fact that \citet{Marthi2014} make use of the ADI method is not immediately clear
from the paper itself. We will elaborate on this in Section~\ref{sec:adi}. The second is to use the preconditioned conjugate gradient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem 
to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before.
Therefore, we also compare the ADI \citep{Marthi2014} and PCG \citep{Liu2010} methods in this paper. We decided to focus on the complex framework proposed by \citet{Smirnov2015} rather than the traditional phase-and-amplitude approach as it 
allows us to trivially prove that the ADI method proposed by \citep{Marthi2014} is merely a special case of the more general LM algorithm.

%Moreover, this framework also 
%allows us to show that the ADI method proposed by \citep{Marthi2014} can for all intents and purposes be renamed to redundant \textsc{StEfCal} \citep{Mitchell:MWA-cal,Salvini2014}.

In Section~\ref{sec:sky_wirtinger}, we review the skymodel-based complex-and-conjugate calibration approach proposed by \citet{Smirnov2015}.
In Section~\ref{sec:red_wirtinger}, we show that this same approach can also be used to perform redundant calibration.
In Section~\ref{sec:results}, we compare the computational complexity of two recently proposed redundant calibration speedup algorithms with one another.
We end the paper with some final conclusions.

@@

{\bf GB: ok here are my comments/questions.
\begin{itemize}
\item I think we wanna start with Section 2 being a short review of standard calibration but in the complex optimization framework. Therefore I would begin with section 3.1, following by section 2.2. In particular I think that the discussion of least squares should be connected to (current) eq. 15;
\item we need more context for section 3.2. Why do we need to introduce a different indexing? This should be motivated. As it stands, I think it should remain where it is if we can make a connection to standard calibration, otherwise it should go in section 3.3;
\item section 3.3 should merge with 4 where redundant calibration is introduced;
\item section 3.5 goes together with the description of the simulations;
\item section 4.1: I do not understand the definition in eq. 32. How is the ratio of two matrices defined? I also suggest to use the symbol $\equiv$ rather than $=$ for definitions throughout the paper;
\item can section 4.2.2 be moved to the appendix? Or how much of it?
\item section 5 should be called something like "Implementation"... In Oleg's paper he writes "Inverting $J^H J$ and separability"... I find it a little too direct, but we need to find something else as these are not yet "Results". I also suggest to revert the order in the way this is presented: we first show that the Hessian is almost diagonal (is this only true in the complex represntation?), then we derive ADI and show that is the same as Marthi \& Chengalur;
\item I have not seen the simulations where you inserted a phase wrap and you calibrate it. Is there a reason why it is not included?
\end{itemize}
}

% \subsection{Least Squares}
% \label{sec:ls}
% The method of nonlinear least squares is tantamount to solving the following optimization problem:  
% \begin{equation}
% \label{eq:least_squares}
% \min_{\bzeta} \Lambda(\bzeta) = \min_{\bzeta} \|\brho\|_F^2 = \min_{\bzeta} \|\bdelta - \bnu(\bzeta)\|_F^2, 
% \end{equation}
% where $\Lambda$ is known as the objective function, $\brho$ is the residual vector, $\bdelta$ is the observed data vector, $\bnu$ is the model-predicted data vector and $\bzeta$ is the model-parameter vector that minimizes Eq.~\eqref{eq:least_squares}.
% In Eq.~\eqref{eq:least_squares}, we assume that the optimization problem is real valued.
% The main objective of least squares is to find the model-parameters which minimize the squared difference between the observed data and the model-predicted data.
% 
% The most standard way of solving Eq.~\eqref{eq:least_squares} is to use a gradient-type minimization algorithm. In this paper we will be focusing on the following gradient minimization algorithms: GN and LM \citep{Levenberg1944,Marquardt1963}. 
% The GN update step is defined as
% \begin{equation}
%  \Delta \bzeta = (\bJ^T\bJ)^{-1} \bJ^T\brho,
% \end{equation}
% while the LM update takes the following form
% \begin{equation}
%  \Delta \bzeta = (\bJ^T\bJ+\lambda\bD)^{-1} \bJ^T\brho,
% \end{equation}
% where $\bJ$ is the Jacobian matrix (i.e. the derivative of the model $\bnu$ with respect to the parameter vector $\bzeta$), $\lambda$ denotes the 
% damping factor and $\bD$ is a diagonal matrix whose diagonal entries are equal to the diagonal entries of $\bJ^T\bJ$. In this paper, we denote 
% matrix transposition with $(*)^T$.
% These update steps can be used in an iterative manner to determine new estimates of $\bzeta$.
% 
% The gradient based minimization algorithms that are generally used to solve least squares problems (i.e. GN and LM) generally require the model $\bnu$ to be differentiable
% towards each model parameter. When the least squares problem is complex it becomes less straightforward to apply these gradient based minimization methods, as
% many complex functions are not analytic (no Taylor expansion around a fiducial point exists) and therefore not differentiable if the classic notion of differentiation is used. For instance, if the 
% classical definition of differentiation is used then $\frac{\partial \conj{z}}{\partial z}$, where $z \in \mathbb{C}$, does not exist.
% 
% The only way to circumvent the differentiability conundrum associated with complex least square problems is to recast our complex optimization problem in such a way that $\bnu$ becomes analytic in its argument.
% The standard way of achieving this is to divide the complex optimization problem into its real and imaginary parts and then to solve for the real and imaginary parts of the parameters separately.
% 
% Recently, \citet{Sorber2012} proposed an alternative strategy, which involves the use of Wirtinger calculus \citep{Wirtinger1927}. The Wirtinger derivatives 
% are defined to be 
% \begin{align}
% \label{eq:wir}
% \frac{\partial}{\partial z} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} -  i \frac{\partial}{\partial y} \right ),&\frac{\partial}{\partial \conj{z}} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} +  i \frac{\partial}{\partial y} \right ). 
% \end{align}
% Using the above definitions we can now easily compute the following:
% \begin{align}
% \frac{\partial z}{\partial z} & = 1, & \frac{\partial \conj{z}}{\partial z}&=0, & \frac{\partial z}{\partial \conj{z}} & = 0, & \frac{\partial \conj{z}}{\partial \conj{z}}&=1.
% \end{align}
% In this alternative approach we treat the complex parameters and their conjugates as separate variables.

\section{Skymodel-based Wirtinger Calibration}
\label{sec:sky_wirtinger}
%In Section~\ref{sec:s_mes}, we present the scalar measurement equation. Next, we define some indexing functions in Section~\ref{sec:indexing}. The measurement 
%equation and the indexing functions from Section~\ref{sec:indexing} are then used in Section~\ref{sec:r_mes} to phrase redundant calibration as a least-squares problem, i.e. Eq.~\eqref{eq:least_squares_red}.
%We present the signal-to-noise ratio (SNR) definition we used in this paper in Section~\ref{sec:snr}. The limitations of this paper are discussed in Section~\ref{sec:scope}.

%\subsection{Scalar Measurement Equation}
%\label{sec:s_mes}
The observed visibility $d_{pq}$ measured by the baseline formed by antenna $p$ and $q$ is described by
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q}y_{pq} + n_{pq},
\end{equation}
where $g_{p}$ and $g_{q}$ are the direction independent gains associated with antenna $p$ and $q$, $y_{pq}$ denotes the true visibility that baseline $pq$ measured
and $n_{pq}$ is the thermal noise component. Conjugation is denoted by $\conj{(*)}$. The true value of $g_p$, $g_q$ and $y_{pq}$ are unknown. We estimate the true value of
$g_p$, $g_q$ and $y_{pq}$ by calibrating our data and then imaging the result.

The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a standard deviation $\sigma$, where   
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}}. 
\end{equation}
Here, $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

>>>>

%\title[Complex redundant calibration]{Redundant calibration as a complex optimization problem}
\title[Redundant interferometric calibration as a complex optimization problem]{Redundant calibration as a complex optimization problem}
%\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
%A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
%$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
%$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
%$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
%$^{4}$National Radio Astronomy Obs., Socorro NM\\
%$^{5}$Cavendish Lab., Cambridge UK\\
%$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, C.L.~Carilli$^{3,4}$, J.S.~Kenyon$^{1}$, A.R.~Parsons$^{5,6}$ \newauthor and O.M.~Smirnov$^{1,2}$\\
$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{3}$National Radio Astronomy Obs., Socorro NM\\
$^{4}$Cavendish Lab., Cambridge UK\\
$^{5}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - ''redundant calibration" - and boosts sensitivity on specific scales. We show that redundant calibration can be formulated in the domain of complex functions of a complex variable. This framework reduces the uncertaninties

@@

@@

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. Measurements of the EoR are difficult because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2015,McQuinn16} and the presence of foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2011,Dillon2012,Parsons2014} and that can be separated by leveraging only on the different spectral properties (citations needed!!!). An exquisite calibration is therefore required as  miscalibrated foreground emission can jeopardize the EoR signal \citep[e.g.,][]{barry2016,grobler2014,ewall-wice2016}. These reasons motivated the design and deploment of arrays in redundant configurations like, MIT-EoR  (reference!) the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Parsons2014,Jacobs2015,Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

Redundant configurations have historically been abandoned because of their poor imaging performances as they measure only a limited number of spatial scales and, therefore, do not adequately sample the sky brightness distribution. For this very reason, however, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio modes for EoR measurements \citep[][]{Parsons2012,Dillon2015}.

The second advantage of redundant array configurations is that they, by definition, they measure the same spatial scales {\bf (GB: perhaps worth having the Van cittert-Zernike theorem here?)} regardless of what the sky brightness distribution looks like. This means that the corruptions that are introduces by the instrument can be calibrated - up to some extent - without knowing anything about the sky brightness distribution \citep[][]{Noordam1982,Wieringa1992}. This property is particularly desirable for EoR observations as it mitigates errors introduced by an incomplete knowledge of the sky brightness distribution.

In this paper we extend the stefcal calibration algorithm to the redundant use case using the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
We also perform a proper computational analysis of the recently proposed (preconditioned conjugate gradient) PCG method when it is applied to the redundant calibration problem.
We also compare these two approaches.

present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. With respect to current algorithms, it is more robust to initial conditions, it is comparatively fast and {\bf ?????...}. We also present an analysis of its computational complexity showing that the LM algorithm shows better convergence properties than the GN currently used \cite{Zheng14}.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015,Kurien2016} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we test the algorithm performances in Section~\ref{sec:results} and we conclude in Section\ref{sec:conclusions}.
%The effects (i.e. the ionosphere, antenna gains, etc.) a celestial signal experiences along its propagation path are modelled with the radio interferometric measurement equation (RIME) \citep{ME1,RRIME1}. Calibration is the procedure by which we remove the errors that these propagation effects introduce into interferometric data and is normally accomplished using a least-squares solver. In the case where we only model the antenna gains, which is the most basic form of the RIME, we can define calibration as finding the antenna gains that minimize the difference between our observed and predicted visibilities. 
%There are two recent developments in skymodel-based calibration which should be explicitly mentioned at this point, since they form the foundation this paper builds on. The first is the discovery of \textsc{StEfCal} (Statistically Efficient Calibration) \citep{Mitchell:MWA-cal,Salvini2014}, which is an alternating direction implicit (ADI) skymodel-based calibration algorithm. It works by assuming the gains are known while solving for their conjugates and then performing the reverse. 
%This procedure results in the linearization of the calibration algorithm and reduces the computational complexity of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$, where $N$ denotes the number of antennas in the array. 
%The second development is the idea: instead of breaking the calibration problem into its real and imaginary parts and then solving for the real and imaginary parts of the unknown parameters separately we can treat the antenna gains and their conjugates as separate parameters to estimate \citep{Smirnov2015}. \citet{Smirnov2015} also showed that the \textsc{StEfCal} algorithm itself is merely a special case of this gain-and-conjugate framework. 
%Two baselines are redundant if their baseline difference vectors have the same length and orientation, which imply that they measure the same $uv$-modes. When we use a redundant array layout, we dramatically reduce the number of unknowns. If our array is redundant enough we reduce the number of unknowns to such an extent that we are able to solve for the unknown observed visibilities themselves in addition to the antenna gains. Redundant calibration is the procedure by which we estimate both the unknown observed visibilities and the antenna gains.
%Redundant calibration was popularized in the 1980's. \citet{Noordam1982} used redundant calibration and the WSRT (Westerbork Synthesis Radio Telescope) to produce high-dynamic range images of 3C84.
%The first implementations of redundant calibration made use of the logarithm function, because when the logarithm is applied to the scalar measurement equation, redundant calibration is transformed into a linear problem \citep{Wieringa1992,Camps2003,Liu2010}. The implementation associated with this approach is commonly referred to as \textsc{logcal}.
%We briefly discuss \textsc{logcal} in Appendix~\ref{sec:logcal}. The next improvement in redundant calibration was to perform a first order Taylor expansion, which is equivalent to the Gauss-Newton (GN) algorithm \citep{Kurien2016}, of the measurement equation and then to solve for the perturbations of the model-parameters \citep{Liu2010}. \citet{Liu2010} perform the first order Taylor expansion of the measurement equation by splitting the unknown parameters into their phase-and-amplitude components. The implementation of  this approach is called \textsc{lincal}. \textsc{lincal} is derived from first principals in Appendix~\ref{sec:lincal}. 
%Currently, redundant calibration is usually implemented using a layered algorithmic strategy. \textsc{logcal} and \textsc{lincal} are both implemented in \textsc{omnical}\footnote{https://github.com/jeffzhen/omnical}, which is a layered redundant calibration pipeline \citep{Zheng2014,Ali2015}.
%The solutions of \textsc{logcal} are used as initial input for \textsc{lincal}, since the GN algorithm requires that its starting parameter guesses be in the vicinity of the true solution.
%It is also interesting to mention here that a completely different calibration strategy can also be used to calibrate redundant arrays, \textsc{CorrCal} \citep{Sievers2017}\footnote{https://github.com/sievers/corrcal2}. In \textsc{CorrCal} the unknown visibilities are no longer solved for. The sky is actually modelled, via two covariance matrices. 
%The first covariance matrix is associated with known point sources. The second covariance matrix models the remaining sky in a statistical sense. The unknown gains, however, are still estimated. We do not elaborate any further on this approach in this paper.
%New instruments like LOFAR (Low Frequency Array) \citep{Noorishad2012}, PAPER (Precision Array for Probing the Epoch of Reionization) \citep{Ali2015} and HERA (Hydrogen Epoch of Reionization Array) \citep{deboer2015} have also sparked a renewed interest in redundant calibration. Using redundancy is especially advantageous when we are trying to detect the EoR (Epoch of Reionization) 21 cm signal, which is inherently quite faint \citep{Parsons2012}. Since redundant baselines make statistically independent measurements of the same $uv$-modes redundancy can greatly improve the SNR (signal-to-noise ratio) of EoR detection experiments. This explains why PAPER and HERA are highly redundant arrays, as they are both primarily used for EoR 21 cm detection experiments.
%\noindent
%\textbf{[GIANNI TO REWRITE AND EXTEND EOR SECTION]}
%In this paper we propose to use the complex framework introduced by \citet{Smirnov2015} instead of the traditional amplitude-and-phase approach used by \textsc{lincal}.
%We also propose to use the Levenberg-Marquardt (LM) gradient-based algorithm instead of the Gauss-Newton (GN) algorithm as it exhibits better convergence properties.
%Using the LM algorithm reduces the number of algorithmic layers that are needed to perform redundant calibration. 
%Two ways of speeding up redundant calibration have recently been proposed. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and the true observed visibilities \citep{Wijnholds2012,Marthi2014}. Note that the fact that \citet{Marthi2014} make use of the ADI method is not immediately clear from the paper itself. We will elaborate on this in Section~\ref{sec:adi}. The second is to use the preconditioned conjugate gradient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before. Therefore, we also compare the ADI \citep{Marthi2014} and PCG \citep{Liu2010} methods in this paper. We decided to focus on the complex framework proposed by \citet{Smirnov2015} rather than the traditional phase-and-amplitude approach as it allows us to trivially prove that the ADI method proposed by \citep{Marthi2014} is merely a special case of the more general LM algorithm. 
%In Section~\ref{sec:sky_wirtinger}, we review the skymodel-based complex-and-conjugate calibration approach proposed by \citet{Smirnov2015}.
%In Section~\ref{sec:red_wirtinger}, we show that this same approach can also be used to perform redundant calibration.
%In Section~\ref{sec:results}, we compare the computational complexity of two recently proposed redundant calibration speedup algorithms with one another.
%We end the paper with some final conclusions.

@@

%{\bf GB: ok here are my comments/questions.
%\begin{itemize}
%\item I think we wanna start with Section 2 being a short review of standard calibration but in the complex optimization framework. Therefore I would begin with section 3.1, following by section 2.2. In particular I think that the discussion of least squares should be connected to (current) eq. 15;
%\item we need more context for section 3.2. Why do we need to introduce a different indexing? This should be motivated. As it stands, I think it should remain where it is if we can make a connection to standard calibration, otherwise it should go in section 3.3;
%\item section 3.3 should merge with 4 where redundant calibration is introduced;
%\item section 3.5 goes together with the description of the simulations;
%\item section 4.1: I do not understand the definition in eq. 32. How is the ratio of two matrices defined? I also suggest to use the symbol $\equiv$ rather than $=$ for definitions throughout the paper;
%\item can section 4.2.2 be moved to the appendix? Or how much of it?
%\item section 5 should be called something like "Implementation"... In Oleg's paper he writes "Inverting $J^H J$ and separability"... I find it a little too direct, but we need to find something else as these are not yet "Results". I also suggest to revert the order in the way this is presented: we first show that the Hessian is almost diagonal (is this only true in the complex represntation?), then we derive ADI and show that is the same as Marthi \& Chengalur;
%\item I have not seen the simulations where you inserted a phase wrap and you calibrate it. Is there a reason why it is not included?
%\end{itemize}
%}

% \subsection{Least Squares}
% \label{sec:ls}
% The method of nonlinear least squares is tantamount to solving the following optimization problem:  
% \begin{equation}
% \label{eq:least_squares}
% \min_{\bzeta} \Lambda(\bzeta) = \min_{\bzeta} \|\brho\|_F^2 = \min_{\bzeta} \|\bdelta - \bnu(\bzeta)\|_F^2, 
% \end{equation}
% where $\Lambda$ is known as the objective function, $\brho$ is the residual vector, $\bdelta$ is the observed data vector, $\bnu$ is the model-predicted data vector and $\bzeta$ is the model-parameter vector that minimizes equation~\ref{eq:least_squares}.
% In equation~\ref{eq:least_squares}, we assume that the optimization problem is real valued.
% The main objective of least squares is to find the model-parameters which minimize the squared difference between the observed data and the model-predicted data.
% 
% The most standard way of solving equation~\ref{eq:least_squares} is to use a gradient-type minimization algorithm. In this paper we will be focusing on the following gradient minimization algorithms: GN and LM \citep{Levenberg1944,Marquardt1963}. 
% The GN update step is defined as
% \begin{equation}
%  \Delta \bzeta = (\bJ^T\bJ)^{-1} \bJ^T\brho,
% \end{equation}
% while the LM update takes the following form
% \begin{equation}
%  \Delta \bzeta = (\bJ^T\bJ+\lambda\bD)^{-1} \bJ^T\brho,
% \end{equation}
% where $\bJ$ is the Jacobian matrix (i.e. the derivative of the model $\bnu$ with respect to the parameter vector $\bzeta$), $\lambda$ denotes the 
% damping factor and $\bD$ is a diagonal matrix whose diagonal entries are equal to the diagonal entries of $\bJ^T\bJ$. In this paper, we denote 
% matrix transposition with $(*)^T$.
% These update steps can be used in an iterative manner to determine new estimates of $\bzeta$.
% 
% The gradient based minimization algorithms that are generally used to solve least squares problems (i.e. GN and LM) generally require the model $\bnu$ to be differentiable
% towards each model parameter. When the least squares problem is complex it becomes less straightforward to apply these gradient based minimization methods, as
% many complex functions are not analytic (no Taylor expansion around a fiducial point exists) and therefore not differentiable if the classic notion of differentiation is used. For instance, if the 
% classical definition of differentiation is used then $\frac{\partial \conj{z}}{\partial z}$, where $z \in \mathbb{C}$, does not exist.
% 
% The only way to circumvent the differentiability conundrum associated with complex least square problems is to recast our complex optimization problem in such a way that $\bnu$ becomes analytic in its argument.
% The standard way of achieving this is to divide the complex optimization problem into its real and imaginary parts and then to solve for the real and imaginary parts of the parameters separately.
% 
% Recently, \citet{Sorber2012} proposed an alternative strategy, which involves the use of Wirtinger calculus \citep{Wirtinger1927}. The Wirtinger derivatives 
% are defined to be 
% \begin{align}
% \label{eq:wir}
% \frac{\partial}{\partial z} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} -  i \frac{\partial}{\partial y} \right ),&\frac{\partial}{\partial \conj{z}} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} +  i \frac{\partial}{\partial y} \right ). 
% \end{align}
% Using the above definitions we can now easily compute the following:
% \begin{align}
% \frac{\partial z}{\partial z} & = 1, & \frac{\partial \conj{z}}{\partial z}&=0, & \frac{\partial z}{\partial \conj{z}} & = 0, & \frac{\partial \conj{z}}{\partial \conj{z}}&=1.
% \end{align}
% In this alternative approach we treat the complex parameters and their conjugates as separate variables.

%\section{Skymodel-based Wirtinger Calibration}
%\section{Calibration described through the Wirtinger formalism}
\section{Wirtinger calibration}
\label{sec:sky_wirtinger}

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is given therefore by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

Considering the number of visibilities $B$ (i.e. baselines) measured by an array of $N$ elements $B = \frac{N^2-N}{2}$, equation~\ref{eq:vis_definition} can be expressed in the following vector form:
\begin{equation}
\label{eq:vis_linear_definition}
\bd = \bv + \bn, 
\end{equation}
where 
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{pq} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_linear_definitions}
\end{align}
and 
\begin{equation}
\alpha_{pq} =
\begin{cases}
(q-p) + (p-1)\left (N-\frac{1}{2}p \right ) & \textrm{if}~p<q\\
%B + (p-q) + (q-1)\left (N-\frac{1}{2}q \right )) & \textrm{if}~p>q\\
%B+q + \frac{1}{2}(p-1)(p-2) & \textrm{if}~p>q\\
0 & \textrm{otherwise}
\end{cases}.
\end{equation}
The function $\alpha_{pq}$ therefore maps composite antenna indexes to unique single indexes, i.e:
\begin{equation}
\{\alpha_{12},\alpha_{13},\cdots,\alpha_{N-1N}\} = \{1,2,\cdots,B\} 
\end{equation}
%\begin{eqnarray}
%\{1,2,\cdots,N-1,N,\cdots,B\}	 =  \nonumber\\
%				 =  \{\alpha_{12},\alpha_{13},\cdots,\alpha_{1N},\alpha_{23},\cdots,\alpha_{N-1N}\} 
%\end{eqnarray}
%and 
%\begin{equation}
%\{1,2,\cdots,N-1,N,\cdots,B\}
%\end{equation}
%are equal. We denote the number of baselines in an array with $B$, i.e. $B = \frac{N^2-N}{2}$. 
The vectors in equation~\ref{eq:vec_linear_definitions} are column vectors of size $B$ (i.e. $p<q$).

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function and $\br$ is the residual vector. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources. The knowledge of $y_{pq}$ is often named ``sky model".
